---
title: "Dimensionality reduction"
author: "Valentina Zangirolami"
date: "2025-05-16"
output: html_document
---

Load libraries

```{r, include=TRUE, warning=FALSE, message=FALSE}
library(tidymodels)
library(ggplot2)
library(NMF)
library(tidyverse)
library(h2o)
```

### **Android Dataset**

```{r, include=TRUE, warning=FALSE, message=FALSE}
android_data <- read.csv("android_data.csv")
head(android_data)

summary(android_data)

#split target and features
label_android <- android_data |> select(Label)
android_data <- android_data |> select(-Label)

```

The final goal: Detect android malware

#### **PCA**

```{r, include=TRUE, warning=FALSE, message=FALSE}
pca_recipe <- recipe(~ ., data = android_data) |>
  step_pca(all_numeric(), num_comp = 84)  # All 40 PCs for evaluation

pca_prepped <- prep(pca_recipe, android_data)
pca_variance <- tidy(pca_prepped, 1, type = "variance")

# Automated selection: Find smallest #PCs reaching 95% variance
optimal_pcs_95 <- pca_variance |>
  filter(terms == "cumulative percent variance") |>
  filter(value >= 95) |>
  slice(1) |>
  pull(component)

optimal_pcs_90 <- pca_variance |>
  filter(terms == "cumulative percent variance") |>
  filter(value >= 90) |>
  slice(1) |>
  pull(component)

optimal_pcs_85 <- pca_variance |>
  filter(terms == "cumulative percent variance") |>
  filter(value >= 85) |>
  slice(1) |>
  pull(component)

cat("Optimal #PCs:", optimal_pcs_85, "(85% variance explained)")
cat("Optimal #PCs:", optimal_pcs_90, "(90% variance explained)")
cat("Optimal #PCs:", optimal_pcs_95, "(95% variance explained)")
```

Fixing different percentage of explained variance, we can obtain the optimal number of components.

```{r, include=TRUE, warning=FALSE, message=FALSE}
pca_variance |>
  filter(terms == "percent variance") |>
  ggplot(aes(as.numeric(component), value)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = optimal_pcs_95, linetype = "dashed", color = "red") +
  geom_vline(xintercept = optimal_pcs_90, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = optimal_pcs_85, linetype = "dashed", color = "violet") +
  labs(title = "PCA Scree Plot", x = "Principal Component", y = "% Variance")
```

We can observe that we can obtain good percentage of explained variance with few components.

```{r, include=TRUE, warning=FALSE, message=FALSE}
pca_recipe_90 <- recipe(~ ., data =android_data ) %>%
  step_normalize(all_numeric()) %>%     # Center/scale numeric vars
  step_pca(all_numeric(), num_comp = 3)

pca_recipe_95 <- recipe(~ ., data =android_data) %>%
  step_normalize(all_numeric()) %>%     # Center/scale numeric vars
  step_pca(all_numeric(), num_comp = 5)
```

#### **NMF**

It is based on 

$$V \approx W H$$
approximating the feature matrix with W and H.

- $W$ is the "basis" matrix with dimension $N \times K$
- $H$ is the "coefficients" matrix with dimension $K \times P$

where N is the number of observations, K is the number of latent variables and P is the number of features.

```{r, include=TRUE, warning=FALSE, message=FALSE}
android_numeric <- android_data |>
  select_if(is.numeric)  # Keeps only numeric columns 

#rescale to obtain positive entries
android_nonneg <- android_numeric |> 
  mutate(across(everything(), ~ .x - min(.x, na.rm = TRUE)))
```


```{r, include=TRUE, warning=FALSE, message=FALSE}
nmf_results <- nmf(as.matrix(android_nonneg), rank = 2:8, method = "lee", nrun = 10)

# Plot metrics (higher cophenetic = better stability)
plot(nmf_results, main = "NMF Rank Selection")
```

Interpretation:

 - **Cophenetic** is a correlation measure about the pairwise distance between points in the original dataand the pairwise distance between points in the reduced data (high cophenetic --> distances are better preserved)
 - **Dispersion**:  stability of NMF across runs
 - **Evar**: variance captured by different ranks
 - **Residuals**: based on V - WH
 - **rss**: residual sum of squared 
 - **Silhouette**: it measures how well each sample fits its assigned cluster compared to other clusters (then, in terms of cohesion and separation
 - **Sparseness**: sparsity in the observations (basis) or in the covariates (coefficients) for each run

A good choice can be rank=5 since we have high cophenetic, low rss and residuals, the sparseness is somehow controlled, the explain variance is high.

Check dimension of W:

```{r, include=TRUE, warning=FALSE, message=FALSE}
nmf_result <- nmf(as.matrix(android_nonneg), rank = 5, method = "lee", nrun = 10)

V.hat <- fitted(nmf_result)
w <- basis(nmf_result) #  W 
dim(w)
```

Check dimension of H:

```{r, include=TRUE, warning=FALSE, message=FALSE}
h_matrix <- coef(nmf_result) #  H 
dim(h_matrix)
```

#### **Autoencoder**

```{r, include=TRUE, warning=FALSE, message=FALSE}
h2o.init()

# Convert data to H2O frame
data_h2o <- as.h2o(android_data)

# Train autoencoder
autoencoder <- h2o.deeplearning(
  x = names(data_h2o),
  training_frame = data_h2o,
  autoencoder = TRUE,
  hidden = c(32, 8, 32),
  activation = "RectifierWithDropout",
  l1 = 1e-5,
  l2 = 1e-5,
  max_w2 = 10,
  adaptive_rate = TRUE,
  input_dropout_ratio = 0.1,
  hidden_dropout_ratios = c(0.1, 0.1, 0.1),
  epochs = 100
)

# Extract latent features
latent_rep <- h2o.deepfeatures(autoencoder, data_h2o, layer = 2) # we can extract second layer
latent_rep_df <- as.data.frame(latent_rep)
```


#### **Compare a model with different kind of features**

In this section, we are going to compare a random forest using

(1) the initial features
(2) the component obtained from PCA
(3) the latent variables obtained from NMF
(4) the latent variables obtained from autoencoder

```{r, include=TRUE, warning=FALSE, message=FALSE}
android_true <- android_data |> mutate(Y = as.factor(label_android$Label))
pca_90 <- bake(prep(pca_recipe_90, training = android_data), new_data = android_data)
android_pca90 <- pca_90 |> mutate(Y = as.factor(label_android$Label))
pca_95 <- bake(prep(pca_recipe_95, training = android_data), new_data = android_data)
android_pca95 <- pca_95 |> mutate(Y = as.factor(label_android$Label))
android_nmf <- as.data.frame(w) |> mutate(Y = as.factor(label_android$Label))
android_autoencoder <- latent_rep_df |> mutate(Y = as.factor(label_android$Label))


set.seed(123)  # For reproducibility
split <- initial_split(android_true, prop = 0.8)
train_data <- training(split)
val_data <- testing(split)

# true features
android_recipe <- recipe(Y~., train_data) |> 
  step_zv(all_predictors())

rf_spec <- rand_forest(mtry = 5, trees = 20, min_n = 20) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

# Create workflow
rf_wf <- workflow() |>
  add_recipe(android_recipe) |>
  add_model(rf_spec)

# Fit model
rf_fit <- parsnip::fit(rf_wf, data = train_data)

# Get predictions on validation data
val_preds_true <- predict(rf_fit, new_data = val_data) |> 
  bind_cols(val_data) 
metrics <- metric_set(accuracy, f_meas)

val_metrics_true <- metrics(val_preds_true, truth = Y, estimate = .pred_class)

#PCA 90%

set.seed(123)  # For reproducibility
split <- initial_split(android_pca90, prop = 0.8)
train_data <- training(split)
val_data <- testing(split)

# true features
android_recipe <- recipe(Y~., train_data) |> 
  step_zv(all_predictors())

rf_spec <- rand_forest(mtry = 5, trees = 20, min_n = 20) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

# Create workflow
rf_wf <- workflow() |>
  add_recipe(android_recipe) |>
  add_model(rf_spec)

# Fit model
rf_fit <- parsnip::fit(rf_wf, data = train_data)

# Get predictions on validation data
val_preds_pca90 <- predict(rf_fit, new_data = val_data) |> 
  bind_cols(val_data) 

val_metrics_pca90 <- metrics(val_preds_pca90, truth = Y, estimate = .pred_class)

#PCA 95%

set.seed(123)  # For reproducibility
split <- initial_split(android_pca95, prop = 0.8)
train_data <- training(split)
val_data <- testing(split)

# true features
android_recipe <- recipe(Y~., train_data) |> 
  step_zv(all_predictors())

rf_spec <- rand_forest(mtry = 5, trees = 20, min_n = 20) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

# Create workflow
rf_wf <- workflow() |>
  add_recipe(android_recipe) |>
  add_model(rf_spec)

# Fit model
rf_fit <- parsnip::fit(rf_wf, data = train_data)

# Get predictions on validation data
val_preds_pca95 <- predict(rf_fit, new_data = val_data) |> 
  bind_cols(val_data) 

val_metrics_pca95 <- metrics(val_preds_pca95, truth = Y, estimate = .pred_class)

# NMF

set.seed(123)  # For reproducibility
split <- initial_split(android_nmf, prop = 0.8)
train_data <- training(split)
val_data <- testing(split)

# true features
android_recipe <- recipe(Y~., train_data) |> 
  step_zv(all_predictors())

rf_spec <- rand_forest(mtry = 5, trees = 20, min_n = 20) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

# Create workflow
rf_wf <- workflow() |>
  add_recipe(android_recipe) |>
  add_model(rf_spec)

# Fit model
rf_fit <- parsnip::fit(rf_wf, data = train_data)

# Get predictions on validation data
val_preds_nmf <- predict(rf_fit, new_data = val_data) |> 
  bind_cols(val_data) 

val_metrics_nmf <- metrics(val_preds_nmf, truth = Y, estimate = .pred_class)

# Autoencoder

set.seed(123)  # For reproducibility
split <- initial_split(android_autoencoder, prop = 0.8)
train_data <- training(split)
val_data <- testing(split)

# true features
android_recipe <- recipe(Y~., train_data) |> 
  step_zv(all_predictors())

rf_spec <- rand_forest(mtry = 5, trees = 20, min_n = 20) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

# Create workflow
rf_wf <- workflow() |>
  add_recipe(android_recipe) |>
  add_model(rf_spec)

# Fit model
rf_fit <- parsnip::fit(rf_wf, data = train_data)

# Get predictions on validation data
val_preds_autoencoder <- predict(rf_fit, new_data = val_data) |> 
  bind_cols(val_data) 

val_metrics_autoencoder <- metrics(val_preds_nmf, truth = Y, estimate = .pred_class)

round(val_metrics_true$.estimate,4)
round(val_metrics_pca90$.estimate,4)
round(val_metrics_pca95$.estimate,4)
round(val_metrics_nmf$.estimate,4)
round(val_metrics_autoencoder$.estimate,4)
```

We can observe from the results that we are not loosing lot of performances.
