---
title: "Ensemble methods"
author: "Valentina Zangirolami (valentina.zangirolami@unimib.it)"
date: "2025-05-09"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
#load libraries
library(car)
library(ggplot2)
library(tidymodels)
library(baguette)
library(fastAdaboost)
library(vip)
library(tidyverse)
```

## ***Dataset: ESG DATA***

```{r, include=TRUE}
esg_data <- read.csv("esg_financial_dataset.csv")
head(esg_data)
```

See the description and data analysis in Multiple Linear regression file.

As before, we should split in training and validation sets

```{r}
set.seed(123)  # For reproducibility
split <- initial_split(esg_data, prop = 0.8)

train_data <- training(split)
val_data <- testing(split)
```

#### **Ensemble methods: Bagging**

Bagging implementation is already provided by bag_tree() from *baguette* library.
We try to implement bagging into a regression tree with the same hyperparameters of our first model in decision_trees.R

```{r}
# Create recipe
esg_recipe <- recipe(MarketCap ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors())

# Create bagging model
bag_spec <- bag_tree(cost_complexity = 0.01, tree_depth = 30, min_n = 20) %>%
  set_engine("rpart", times = 50) %>%
  set_mode("regression")

# Create workflow
bag_wf <- workflow() %>%
  add_recipe(esg_recipe) %>%
  add_model(bag_spec)

# Fit model
bag_fit <- fit(bag_wf, data = train_data)

# Get predictions on training data
train_preds <- predict(bag_fit, new_data = train_data) %>% 
  bind_cols(train_data)  # Attach true values
# Get predictions on validation data
val_preds <- predict(bag_fit, new_data = val_data) %>% 
  bind_cols(val_data)  # Attach true values

# Compute metrics (e.g., RMSE, R-squared for regression)
metrics <- metric_set(rmse, rsq)
train_metrics <- metrics(train_preds, truth = MarketCap, estimate = .pred)
val_metrics <- metrics(val_preds, truth = MarketCap, estimate = .pred)

round(train_metrics$.estimate, 4)
round(val_metrics$.estimate,4)
```

With bagging, we improved our results. In terms of R-squared, we obtain higher values and a reduced difference between training and validation sets. Also in terms of rmse, the results are better while the distance among these two measures seem to be large.

**Random Forest**

Random Forest combines decision trees and bagging while also sampling randomly features for each tree. The most popular libraries on R are ranger and RandomForest.. In this case, we will use ranger.

Three hyperparameters:

- *mtry*: an integer for the number of predictors that will be randomly sampled at each split when creating the tree models
- *trees*: an integer for the number of trees contained in the ensemble
- *min_n*: an integer for the minimum number of data points in a node that are required for the node to be split further

```{r}
rf_spec <- rand_forest(mtry = 8, trees = 10, min_n = 8) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

# Create workflow
rf_wf <- workflow() %>%
  add_recipe(esg_recipe) %>%
  add_model(rf_spec)

# Fit model
rf_fit <- fit(rf_wf, data = train_data)

# Get predictions on training data
train_preds <- predict(rf_fit, new_data = train_data) %>% 
  bind_cols(train_data)  # Attach true values
# Get predictions on validation data
val_preds <- predict(rf_fit, new_data = val_data) %>% 
  bind_cols(val_data)  # Attach true values

# Compute metrics (e.g., RMSE, R-squared for regression)
metrics <- metric_set(rmse, rsq)
train_metrics <- metrics(train_preds, truth = MarketCap, estimate = .pred)
val_metrics <- metrics(val_preds, truth = MarketCap, estimate = .pred)

round(train_metrics$.estimate, 4)
round(val_metrics$.estimate,4)
```

The results are promising but we still observe overfitting issues. We need to reduce complexity or we can increase the number of trees

```{r}
rf_spec <- rand_forest(mtry = 8, trees = 100, min_n = 8) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

# Create workflow
rf_wf <- workflow() %>%
  add_recipe(esg_recipe) %>%
  add_model(rf_spec)

# Fit model
rf_fit <- fit(rf_wf, data = train_data)

# Get predictions on training data
train_preds <- predict(rf_fit, new_data = train_data) %>% 
  bind_cols(train_data)  # Attach true values
# Get predictions on validation data
val_preds <- predict(rf_fit, new_data = val_data) %>% 
  bind_cols(val_data)  # Attach true values

# Compute metrics (e.g., RMSE, R-squared for regression)
metrics <- metric_set(rmse, rsq)
train_metrics <- metrics(train_preds, truth = MarketCap, estimate = .pred)
val_metrics <- metrics(val_preds, truth = MarketCap, estimate = .pred)

round(train_metrics$.estimate, 4)
round(val_metrics$.estimate,4)
```

We obtained better results on validation set with a slight improvement of overfitting concerns.

Let's visualize the most important features

```{r}

final_model <- extract_fit_engine(rf_fit)

vip(final_model, geom = "point", horizontal = FALSE) +
  ggtitle("Decision Tree Feature Importance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```


#### **Exercise **

Analyze the dataset london_bike_sharing from e-learning and apply appropriate statistical learning while finding good hyperparameters whether it is possible.

```{r, include=TRUE}
london_data <- read.csv("london_bike_sharing.csv")
head(london_data)
```

```{r variable-description, echo=FALSE, message=FALSE}
library(knitr)

# Create a description dataframe
variable_desc <- data.frame(
  Variable = c(
    "daily_cnt", "t1_max", "t1_min", "t1_median", 
    "t2_max", "t2_min", "t2_median", "weather_code", 
    "is_holiday", "is_weekend", "season", 
    "max_hum", "mean_hum", "min_hum", 
    "max_wind_speed", "mean_wind_speed", "min_wind_speed"
  ),
  Type = c(
    "Numeric", "Numeric", "Numeric", "Numeric",
    "Numeric", "Numeric", "Numeric", "Categorical",
    "Binary", "Binary", "Categorical",
    "Numeric", "Numeric", "Numeric",
    "Numeric", "Numeric", "Numeric"
  ),
  Description = c(
    "Total daily count of bike rentals.",
    "Maximum real temperature (°C) recorded during the day.",
    "Minimum real temperature (°C) recorded during the day.",
    "Median real temperature (°C) for the day.",
    "Maximum 'feels like' temperature (°C) (perceived, considering wind/humidity).",
    "Minimum 'feels like' temperature (°C).",
    "Median 'feels like' temperature (°C).",
    "Most frequent weather condition of the day (mode). Common codes: 1 = Clear, 2 = Scattered clouds, 3 = Broken clouds, etc.",
    "1 if the day is a holiday, 0 otherwise.",
    "1 if the day is a weekend (Saturday/Sunday), 0 otherwise.",
    "Season of the year: 1 = Spring, 2 = Summer, 3 = Fall, 4 = Winter.",
    "Maximum humidity (%) during the day.",
    "Mean humidity (%) for the day.",
    "Minimum humidity (%) during the day.",
    "Maximum wind speed (km/h or m/s) recorded.",
    "Mean wind speed for the day.",
    "Minimum wind speed recorded."
  )
)

# Render as a formatted table
kable(variable_desc, caption = "Table 1: Description of Variables in the Daily Bike Rentals Dataset")
```

