---
title: "Polynomial Regression and Splines"
author: "Valentina Zangirolami (valentina.zangirolami@unimib.it)"
date: "2025-05-07"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
#load libraries
library(car)
library(ggplot2)
library(gam)
library(ISLR2)
library(splines)
library(tidymodels)
library(mgcv)       
library(tidyverse)
```

## ***Dataset: College***

```{r, include=TRUE}
data_college = College

head(data_college)
```

College dataset is contained in ISLR2 library. You can find the description of the variables in R while searching "College" in **Help**. **Our goal**: predict the number of enrollments for each new college.

```{r, include=FALSE}
data_college <- data_college |>
  slice(-which.max(Apps))  
```


```{r, include=TRUE, message=FALSE, warning=FALSE}
#marginal interpretation

(Scatter_Matrix <- GGally::ggpairs(data_college[,c(2:10)], 
                                   title = "Scatter Plot Matrix for College (first 9 variables)", 
                                   axisLabels = "show"))
(Scatter_Matrix_2 <- GGally::ggpairs(data_college[,c(4, 11:18)], 
                                   title = "Scatter Plot Matrix for College (remaining variables)", 
                                   axisLabels = "show"))

```

Marginally, the relationships between Enroll and some covariates (such as Accept, Apps, F.Undergrad) seem to be linear. For other covariates (such as PhD, terminal, perc. Alumni), we can observe a clear non-linear pattern with Enroll. However, the marginal interpretation is not always clear, but, in general, we need to overcome linear relationships between response (Enroll) and covariates.

### **Generalized Additive Models (GAMs)**

GAM allows us to model non-linear relationships when many covariates are involved. 

Before starting, we can split our dataset in the training and validation sets as follows

```{r, include=TRUE}

set.seed(123)  # For reproducibility
split <- initial_split(data_college, prop = 0.8)

train_data <- training(split)
val_data <- testing(split)
```

(Instead of holdout sample, you can use also cross-validation. In the following classes, we shall consider it.)
