---
title: "Polynomial Regression and Splines"
author: "Valentina Zangirolami (valentina.zangirolami@unimib.it)"
date: "2025-05-07"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
#load libraries
library(car)
library(ggplot2)
library(ISLR2)
library(splines)
library(tidymodels)
library(yardstick)
library(mgcv)       
library(tidyverse)
```

## ***Dataset: College***

```{r, include=TRUE}
data_college = College

head(data_college)
```

College dataset is contained in ISLR2 library. You can find the description of the variables in R while searching "College" in **Help**. **Our goal**: predict the number of enrollments for each new college.

```{r, include=FALSE}
data_college <- data_college |>
  slice(-which.max(Apps))  
```


```{r, include=TRUE, message=FALSE, warning=FALSE}
#marginal interpretation

(Scatter_Matrix <- GGally::ggpairs(data_college[,c(2:10)], 
                                   title = "Scatter Plot Matrix for College (first 9 variables)", 
                                   axisLabels = "show"))
(Scatter_Matrix_2 <- GGally::ggpairs(data_college[,c(4, 11:18)], 
                                   title = "Scatter Plot Matrix for College (remaining variables)", 
                                   axisLabels = "show"))

```

Marginally, the relationships between Enroll and some covariates (such as Accept, Apps, F.Undergrad) seem to be linear. For other covariates (such as PhD, terminal, perc. Alumni), we can observe a clear non-linear pattern with Enroll. However, the marginal interpretation is not always clear, but, in general, we need to overcome linear relationships between response (Enroll) and covariates.

### **Generalized Additive Models (GAMs)**

GAM allows us to model non-linear relationships when many covariates are involved. 

Before starting, we can split our dataset in the training and validation sets as follows

```{r, include=TRUE}

set.seed(123)  # For reproducibility
split <- initial_split(data_college, prop = 0.8)

train_data <- training(split)
val_data <- testing(split)
```

(Instead of holdout sample, you can use also cross-validation. In the following classes, we shall consider it.)

Since we are considering *Enroll* as response variable, we need to model the response in an appropriate way. 

If we would like to include natural splines for many covariates, it is not necessary leveraging backfitting algorithm. Indeed, we can include each natural spline in the classical formula of Generalized Linear model.

Generally, we assume $Enroll \sim Pois(\mu)$ where

$$log(\mu) = \beta_0 + \sum_{i=1}^{17} f_i(X^{i})$$
with $X^{(i)}$ meaning the 17 covariates. 

```{r}
gam_1 <- glm(Enroll ~ Apps + Accept + Private + F.Undergrad + ns(Top10perc, 3) + ns(Top25perc,3) + P.Undergrad + ns(Outstate,3) + ns(Room.Board, 4) + ns(Books) + ns(Personal,3) + ns(PhD,3) + ns(Terminal,3) + ns(S.F.Ratio,3) + ns(perc.alumni,3) + ns(Expend,3) + ns(Grad.Rate,3),family=poisson, data=train_data)

summary(gam_1)
```

Specifically, the estimated model corresponds to

$$log(\mu_i) = 6.01 + 0.00004 Apps_i + 0.00009 Accept_i -0.33 D_i + 0.00003 F.Undergrad + 0.28 b_1(Top10perc) + 0.57b_2(Top10perc) + 0.47 b_3(Top10perc) + 0.03 b_4(Top25perc) -0.42b_5(Top25perc) - 0.47 b_6(Top25perc) + \ldots,$$ 

where $b_1, b_2, b_3, b_4, b_5, b_6$ are basis functions. 

**Comments**

- Most of coefficients are significant at any $\alpha$ level (except to ns(Top25perc, 3)1, ns(Terminal, 3)2, ns(perc.alumni, 3)3 )
- The test about overall significance:

```{r}
mod_0 <- glm(Enroll ~ 1, family=poisson, data=train_data)
anova(mod_0,gam_1,test="Chisq")
```

The hypothesis of non-significance of all coefficients is rejected.

When we want to include splies as smoothing splines, we need to use backfitting algorithm by using the function *gam*, as follows

```{r, include=TRUE}
gam_2 <- gam(Enroll ~ Apps + Accept + Private + F.Undergrad + 
             s(Top10perc, k=3) + s(Top25perc, k=3) + P.Undergrad + 
             s(Outstate, k=3) + s(Room.Board, k=4) + s(Books, k=3) + 
             s(Personal, k=3) + s(PhD, k=3) + s(Terminal, k=3) + 
             s(S.F.Ratio, k=3) + s(perc.alumni, k=3) + s(Expend, k=3) + 
             s(Grad.Rate, k=3), 
             family = poisson(), 
             data = train_data,
             method = "REML")

summary(gam_2)
```

All the coefficients are significant and the R2 adj is really high. This means that the model is good and it is likely to fit well data.

**Comparison of predictions**

```{r}
pred_gam_1_train <- predict(gam_1, train_data[,-4])
pred_gam_2_train <- predict(gam_2, train_data[,-4])

pred_gam_1 <- predict(gam_1, val_data[,-4])
pred_gam_2 <- predict(gam_2, val_data[,-4])

rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}
cat("RMSE on training data \n")
(rmse_gam_1_train <- rmse(train_data$Enroll, pred_gam_1_train))
(rmse_gam_2_train <- rmse(train_data$Enroll, pred_gam_2_train))
cat("RMSE on validation data \n")
(rmse_gam_1 <- rmse(val_data$Enroll, pred_gam_1))
(rmse_gam_2 <- rmse(val_data$Enroll, pred_gam_2))


```

The RMSE are similar between the two models. We can see that they are better on the training data, which highlights overfitting.